---
title: "Overview of Statistical Learning and Modeling -- 36-600"
author: Trishla Nair
subtitle: |
  Lab 11T: Nonlinear Regression\
  Due Wednesday, November 12, 11:59pm
output:
  pdf_document:
    toc: false
  html_document:
    toc: false
    code-overflow: wrap
    code-fold: false
    theme: spacelab
  word_document:
    toc: false
urlcolor: blue
---

```{r}
suppressMessages(library(tidyverse))
```

## Data

Let's start by simulating a dataset from a nonlinear curve:

```{r}
set.seed(555)
x <- -5:5
y <- 0.1*x^3 - 0.5*x + 2.1 + rnorm(length(x), mean = 0, sd = 0.5*(1 + abs(x)))
e <- 0.5*(1 + abs(x))

df <- data.frame("x" = x, "y" = y, "e" = e)

suppressMessages(library(tidyverse))
ggplot(data = df, mapping = aes(x = x, y = y)) +
  geom_errorbar(aes(ymin = y-e, ymax = y+e), width = 0.1, color = "blue") +
  geom_point(color = "firebrick")
```

With this data, you'll be able to see exactly how different methods perform differently on the same comprehensible dataset.


# Questions


## Question 1

Implement a (weighted) global cubic polynomial regression model in a similar manner to that implemented in the notes; more specifically, this means: learn the model, run `predict()` to determine the regression line, plot the data with the regression line superimposed, show the coefficients, and compute the mean-squared error.  Just like the notes, don't worry about splitting the data into training and test sets.

```{r}
pr.out  <- lm(y ~ poly(x, 3, raw = TRUE), data = df,
              weights = 1/(e^2))
pr.pred <- predict(pr.out)
mean((y - pr.pred)^2)
data.frame("Estimate" = 
             summary(pr.out)$coefficients[, 1])

ggplot(data = df, mapping = aes(x = x, y = y)) +
  geom_errorbar(aes(ymin = y-e, ymax = y+e), width = 0.1, color = "blue") +
  geom_point(color = "firebrick") +
  geom_line(data = data.frame(x, "y" = pr.pred), 
            mapping = aes(x = x, y = y), col = "sienna")
```


## Question 2

Repeat Question 1, but using a regression splines model.  Assume four degrees of freedom.

```{r}
library(splines)
# Default: cubic spline -> K+2 basis functions
#  remember: K is the # of non-overlapping segments
s.out  <- lm(y ~ bs(x, df = 4), data = df,weights = 1/(e^2))
s.pred <- predict(s.out)
mean((y - s.pred)^2)
data.frame("Estimate" = 
             summary(s.out)$coefficients[, 1])

ggplot(data = df, mapping = aes(x = x, y = y)) +
  geom_errorbar(aes(ymin = y-e, ymax = y+e), width = 0.1, color = "blue") +
  geom_point(color = "firebrick") +
  geom_line(data = data.frame(x, "y" = s.pred), 
            mapping = aes(x = x, y = y), col = "sienna")
```


## Question 3

Repeat Question 1, but with a smoothing spline model.  Note that you may get a "surprising" result.

```{r}
ss.out  <- suppressWarnings(
  smooth.spline(df$x, y = df$y, 
                w = 1/(df$e^2), cv = TRUE)
)
ss.out$lambda
ss.pred <- predict(ss.out)
mean((y - ss.pred$y)^2)

ggplot(data = df, mapping = aes(x = x, y = y)) +
  geom_errorbar(aes(ymin = y-e, ymax = y+e), width = 0.1, color = "blue") +
  geom_point(color = "firebrick") +
  geom_line(data = data.frame(x, "y" = ss.pred$y), 
            mapping = aes(x = x, y = y), col = "sienna")
```


## Question 4

Repeat Question 1, but with a local polynomial regression model.  Assume a `span` of 0.6.

```{r}
lpr.out  <- loess(y ~ x, data = df, 
                  weights = 1/(df$e)^2, span = 0.6)
lpr.pred <- predict(lpr.out)
mean((y - lpr.pred)^2)

ggplot(data = df, mapping = aes(x = x, y = y)) +
  geom_errorbar(aes(ymin = y-e, ymax = y+e), width = 0.1, color = "blue") +
  geom_point(color = "firebrick") +
  geom_line(data = data.frame(x, "y" = lpr.pred), 
            mapping = aes(x = x, y = y), col = "sienna")
```


## Question 5

Reform the plot from Question 4, but add a one-standard-error confidence band.  You can do this by running the first line, then subsequently adding the last two lines onto your `ggplot()` call, once they're specified as you prefer:

```
p <- predict(lpr.out, se = TRUE)

+ geom_line(mapping = aes(x = lpr.out$x, y = lpr.out$fitted + p$se), 
            color = "[your color]", linetype = "dashed")
+ geom_line(mapping = aes(x = lpr.out$x, y = lpr.out$fitted - p$se), 
            color = "[your color]", linetype = "dashed")
```

What does the band actually mean?  Because it's a one-standard-error band, it means that for any given $x$, there is an approximately 68% chance that the band overlaps the true underlying function value.  This is generally a rough statement, though, given that *there is almost always correlation between neighboring data points* (i.e., the lack of independence between $y_{i-1}$, $y_i$, and $y_{i+1}$, etc.).  Just think of the band as a notion of how uncertain your fitted curve is at each $x$: is the band wide or narrow?  Note that the bands get wider as we get to either end of the data: this is an expected feature, not a bug.  There's fewer data within the span at either end, so the fitted function is that much more uncertain towards the periphery.

```{r}
lpr.out  <- loess(y ~ x, data = df, weights = 1/(df$e)^2, span = 0.6)
p <- predict(lpr.out, se = TRUE)
pred.df <- data.frame(
  x = lpr.out$x,
  fit = lpr.out$fitted,
  upper = lpr.out$fitted + p$se,
  lower = lpr.out$fitted - p$se
)

ggplot(data = df, aes(x = x, y = y)) +
  geom_errorbar(aes(ymin = y - e, ymax = y + e), width = 0.1, color = "blue") +
  geom_point(color = "firebrick") +
  geom_line(data = pred.df, aes(x = x, y = fit), color = "sienna", size = 1) +
  geom_line(data = pred.df, aes(x = x, y = upper), color = "gray40", linetype = "dashed") +
  geom_line(data = pred.df, aes(x = x, y = lower), color = "gray40", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "LOESS Fit with One-Standard-Error Confidence Band",
    y = "y",
    x = "x"
  )

```


