---
title: Overview of Statistical Learning and Modeling -- 36-600
subtitle: |
  Lab 4R: Unsupervised Learning II\
  Due Friday, September 19, 11:59pm\
author: Trishla Nair
output:
  html_document:
    toc: false
    code-overflow: wrap
    code-fold: false
    theme: spacelab
  pdf_document:
    toc: false
urlcolor: blue
---

This lab will be a relatively short one, given that you are also working through your EDA Project.

To answer the questions below, it again may help you to refer to Sections 10.3 and 10.5 of the ISLR book (the James et al. book listed on the syllabus), as well as previous labs.


## Question 1

Like we did last time, let's create a fake dataset:

```{r}
set.seed(505)
(df <- data.frame(x = runif(3), y = runif(3), z = runif(3)))
```

Compute and show the pairwise distance matrix for the *scaled* data.  (This is what you would pass into `hclust()`, so there is a good reason for doing this.)  You should see a lower-triangular matrix as output, with three values between 1 and 4.  Mentally note the smallest value.

```{r}
d <- dist(scale(df))
d.mat <- as.matrix(d)
hc.out <- hclust(d)
plot(hc.out, main = "Dendrogram (Complete Linkage)")
```


## Question 2

Now run hierarchical clustering, inputting the distance matrix you created above, using *average linkage*, and plot the resulting dendrogram.  What you *should* see is that the height of the first merge is (visually, roughly) the same as that smallest distance value from the last question.  You can confirm this by looking at the `height` element of the list output by `hclust`.  This makes sense: the smallest dissimilarity between our data points is exactly the smallest observed Euclidean distance between the points.

```{r}
d <- dist(scale(df))
hc.out <- hclust(d, method = "average")

plot(hc.out, main = "Dendrogram (Average Linkage)", xlab = "", sub = "")

min_dist <- min(d)
first_merge_height <- hc.out$height[1]
cat("Smallest pairwise distance:", min_dist, "\n")
cat("Height of first merge:", first_merge_height, "\n")

```


---

## Dataset

Let's import the same stellar dataset we used in the previous lab.

```{r}
file.path <- "https://github.com/FSKoerner/F25-36600-data/raw/refs/heads/main/draco_photometry.Rdata"
load(url(file.path))
df <- data.frame(ra, dec, "v" = velocity.los, log.g, 
                 "g" = mag.g, "r" = mag.r, "i" = mag.i)
rm(file.path, ra, dec, velocity.los, log.g, temperature, 
   mag.u, mag.g, mag.r, mag.i, mag.z, metallicity, signal.noise)
suppressMessages(library(tidyverse))
df %>% 
  filter(., ra < 264 & dec > 56 & v > -350 & v < -250) %>% 
  mutate(., gr = g-r, ri = r-i) %>% 
  select(., -g, -r, -i, -v) -> df.new
```


---

## Question 3

Use the `hclust()` function to build a hierarchical clustering tree for `df.new`, and use the basic `plot()` function to display the dendrogram.  Try both complete and average linkage: which one makes for the best-looking output?  (This should **not** be confused with which one gives the best clustering result.  *Note*: there is no "right" answer here; best-looking is in the eye of the statistical consultant.)  Despite talking up the dendrogram in class, is this output actually useful?  Why or why not?  If your client asked for a dendrogram, what step might you want to consider taking before providing one?  (*Note*: when calling `plot()`, consider passing the argument `labels = FALSE` to remove the row numbers at the base of dendrogram.)

```{r}
df.new.scaled <- scale(df.new)

# Compute distances
d.new <- dist(df.new.scaled)

# Complete linkage
hc.complete <- hclust(d.new, method = "complete")
plot(hc.complete, labels = FALSE, main = "Complete Linkage")

# Average linkage
hc.average <- hclust(d.new, method = "average")
plot(hc.average, labels = FALSE, main = "Average Linkage")

```

```
While they show the clusters, a denedrogram is more useful when the data size is smaller because it is more cleaner to understand.
```


## Question 4

Use the `cutree()` function to map each observation to a cluster, then use `ggpairs()` to display the clusters in a similar manner as above for $k$-means.  Assume the same number of clusters as you did for $k$-means.  Does the output look the same or different from that which $k$-means produced?  Is this what you expected?  Why or why not?  (*Hint*: if `cluster` is the output from `cutree()`, then `color = factor(cluster)` will properly color each of the points.)  Visualizing the output of hierarchical clustering in this manner (rather than using a dendrogram) is often preferable as the sample size gets large.

```{r}
library(GGally)
cluster.hc <- cutree(hc.complete, k = 2)
df.plot <- as.data.frame(df.new)
df.plot$cluster <- factor(cluster.hc)
ggpairs(df.plot, aes(color = cluster))
```

```
The ggpairs plot shows that the hierarchical clustering solution looks somewhat similar to the $k$-means output but with noticeable differences in certain regions of the parameter space.
```


## Question 5

Implement a GMM-based analysis using the `ClusterR` package, analogously to what was done in the notes.  Assume *two* clusters.  Your final goal is to figure out the proportions of the observations that can be confidently placed in either Cluster 1 or Cluster 2 (i.e., having cluster probabilities &gt;0.95).  The placement of the rest of the observations can be considered ambiguous.  As a reminder, one of the outputs from `predict_GMM` is `cluster_proba`.  Here, that will be a 1218 x 2 matrix, where the probabilities within each row sum to 1.  So, determine how many values in the first column of `cluster_proba` are either &lt;0.05 (the datum is to be associated with *Cluster 2* with high probability) or &gt;0.95, then divide by the number of rows in `cluster_proba`.  (Note that I found $\approx$ 77.9% of the data can be confidently placed in one of the two clusters.)

(*Note*: you will have to install `ClusterR` and uncomment the line below before completing this question.)
```{r}
library(ClusterR)
gmm.out <- GMM(df.new, gaussian_comps = 2)
pred    <- predict_GMM(df.new, gmm.out$centroids, gmm.out$covariance_matrices, gmm.out$weights)
gmm.out$centroids
names(pred)
```

