---
title: Overview of Statistical Learning and Modeling -- 36-600
subtitle: |
  Lab 14T: Data Analysis Workflow Review\
  Due Wednesday, December 3, 11:59pm
author: "Trishla Nair"
output:
  pdf_document:
    toc: false
  html_document:
    toc: false
    code-overflow: wrap
    code-fold: false
    theme: spacelab
urlcolor: blue
---

This week's lab is very much DIY.  There is a file called `creditDefault.csv` on the course Github page.  Your boss wants to know if any of the variables included in this file, apart from `Default` itself, are important for predicting whether a person will default on their credit-card balance rather than paying it back.

From just the information above, you should eventually construct an appropriate answer to the question.  Note that regardless of whether you think a non-ML model is sufficient, you should always check to see if an ML model gives much better predictive results.  Also note that your boss is not necessarily looking for plots here, but you'd always want to do *some* EDA, if for no other reason than to ensure you don't have missing data or outliers!

```{r}
df <- read.csv("https://raw.githubusercontent.com/FSKoerner/F25-36600-data/refs/heads/main/creditDefault.csv", stringsAsFactors = TRUE)

```

```{r}
boxplot(df)
```

```{r}
hist(df$Balance)
```

```{r}
hist(df$Income)
```

```{r}
set.seed(42)
s        <- sample(nrow(df), round(0.7 * nrow(df)))
df.train <- df[s, ]
df.test  <- df[-s, ]

logit.out <- glm(Default ~ ., data = df.train, family = binomial)
summary(logit.out)
```
```{r}
balance.grid <- seq(min(df.train$Balance), max(df.train$Balance), length = 300)

# Predict for non-student
pred.non <- predict(logit.out,
                    newdata = data.frame(
                      Student = "No",
                      Balance = balance.grid,
                      Income = mean(df.train$Income)
                    ),
                    type = "response")

# Predict for student
pred.stud <- predict(logit.out,
                     newdata = data.frame(
                       Student = "Yes",
                       Balance = balance.grid,
                       Income = mean(df.train$Income)
                     ),
                     type = "response")

# PLot
plot(balance.grid, pred.non, type = "l", lwd = 2,
     xlab = "Balance", ylab = "Predicted Probability of Default",
     main = "Logistic Regression: Predicted Probability of Default")

lines(balance.grid, pred.stud, lwd = 2, col = "blue")
legend("topleft", legend = c("Non-Student", "Student"),
       col = c("black", "blue"), lwd = 2)
```

Based on our analysis of the logistic regression model, we can see that Balance is significant in determining if a someone is likely to default.