---
title: Overview of Statistical Learning and Modeling -- 36-600
subtitle: |
  Lab 7R: Random Effects\
  Due Friday, October 10, 11:59pm\
author: Trishla Nair
output:
  pdf_document:
    toc: false
  html_document:
    toc: false
    code-overflow: wrap
    code-fold: false
    theme: spacelab
urlcolor: blue
---


Today's lab will be relatively short, so that you can turn it in and head off to enjoy fall break.


## Data

Here we read in the sleep study data you used in Project 1 and we retain three variables: `cohort` and `term_gpa` as predictors, and `TotalSleepTime` (renamed `sleep`) as the response.

```{r}
suppressMessages(library(tidyverse))
df        <- read.csv(url("https://raw.githubusercontent.com/FSKoerner/F25-36600-data/refs/heads/main/cmu-sleep.csv"))
df$cohort <- factor(df$cohort)
df %>% select(., cohort, TotalSleepTime, term_gpa) -> df
names(df)[2] <- "sleep"
summary(df)
```

Note that for this lab, we are not going to split the data into training and test sets.


## Question 1

Let's first assume that we regress `sleep` upon `cohort`.  Since the cohorts that appear in the study are a randomly selected subset of all possible cohorts, learning a random effects model is appropriate (as opposed to a one-way ANOVA model).  Fit an appropriate model using `lmer()` and pass the output to `summary()`.  Display the five intercepts for the five cohorts.  (*Hint*: use one of the functions shown in the notes, specifically one used with the Palmer Penguin data.)  Last, what is the within-cohort correlation?  (This last part involves extracting numbers from the `summary()` output and "doing math" as opposed to making further function calls.)  Is this a weak, medium, or strong correlation?  (If you don't think you have a good sense for that strength, consider using the internet to help you with that judgement - find some examples of the strength of correlation in real data settings that you have a good handle on - and bear in mind that rarely will two tables mapping correlation coefficients to adjectives match.)

```{r}
suppressMessages(library(lme4))

model <- lmer(sleep ~ 1 + (1 | cohort), data = df)
summary(model)

coef(model)$cohort

vc <- as.data.frame(VarCorr(model))
var_cohort <- vc[vc$grp == "cohort", "vcov"]
var_resid <- vc[vc$grp == "Residual", "vcov"]

rho <- var_cohort / (var_cohort + var_resid)
rho
```

```
Given rho is approximately 0.078, we can say that there is a week correlation.
```


## Question 2

Now let's look at `term_gpa`.  Learn a simple linear regression model that has `sleep` as the response variable and `term_gpa` as the predictor variable, and overlay the model on a plot of `sleep` versus `term_gpa`.  Also, determine the AIC value for the model.

```{r fig.align='center',fig.height=4,fig.width=4}
lm_gpa <- lm(sleep ~ term_gpa, data = df)
summary(lm_gpa)
aic_value <- AIC(lm_gpa)
print(aic_value)
ggplot(df, aes(x = term_gpa, y = sleep)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "blue",linewidth = 1) +
  labs(
    title = "Relationship Between Sleep and Term GPA",
    x = "Term GPA",
    y = "Sleep (hours)"
  )
```


## Question 3

Now implement a mixed effects model with `term_gpa` as the fixed effect and `cohort` as the random effect, with random intercepts (but sharing the same slope).  Output the linear model coefficients for each cohort.  (Mentally confirm that indeed, the intercepts are different but the slopes are the same.)  And, as you did in Question 2, record the AIC value.

No plot is necessary, for now.

```{r}
mixed1 <- lmer(sleep ~ term_gpa + (1 | cohort), data = df)
summary(mixed1)
coef(mixed1)$cohort
aic_mixed1 <-AIC(mixed1)
```


## Question 4

Repeat Question 3, now allowing random slopes but the same intercept.

```{r}
mixed2 <- lmer(sleep ~ term_gpa + (0+term_gpa | cohort), data = df)
summary(mixed2)
coef(mixed2)$cohort
aic_mixed2 <-AIC(mixed2)
```


## Question 5

Repeat Question 3, but now with random slopes *and* intercepts.

```{r}
mixed3 <- lmer(sleep ~ term_gpa + (term_gpa | cohort), data = df)
summary(mixed3)
coef(mixed3)$cohort
aic_mixed3 <-AIC(mixed3)
```


## Question 6

Select the best model from among those in Questions 2-5 (i.e., the one with the lowest AIC value).  Plot the data again, and overlay the best model; note that this may require you to make five separate calls to `geom_abline()`.  *Note*: to access the slopes and intercepts, you could use, e.g.,

```
coefs <- coef(lmer.out)$cohort
coefs[1, 1]   # first row, first column -- intercept of first cohort
coefs[3, 2]   # third row, second column -- slope of third cohort
etc.
```

In the end, your plot may not clearly show five separate lines, which is fine, because the `lac1` and `lac2`, and `uw1` and `uw2`, cohorts had very similar data.

```{r fig.align='center',fig.height=4,fig.width=4}
#print(aic_value) 
#print(aic_mixed1)# <- best
#print(aic_mixed2)
#print(aic_mixed3)

best_model <- mixed1
coefs <- coef(best_model)$cohort

p <- ggplot(df, aes(x = term_gpa, y = sleep, color = cohort)) +
  geom_point()

# Add regression lines for each cohort
p <- p + 
  geom_abline(intercept = coefs["lac1", 1], slope = coefs["lac1", 2], color = "red") +
  geom_abline(intercept = coefs["lac2", 1], slope = coefs["lac2", 2], color = "blue") +
  geom_abline(intercept = coefs["uw1", 1], slope = coefs["uw1", 2], color = "green") +
  geom_abline(intercept = coefs["uw2", 1], slope = coefs["uw2", 2], color = "purple") +
  geom_abline(intercept = coefs["other", 1], slope = coefs["other", 2], color = "orange") +
  theme_minimal() +
  labs(title = "Term GPA vs Outcome with Best Model Lines",
       x = "Term GPA",
       y = "sleep")
print(p)
```


