tinytex::install_tinytex()

---
title: Overview of Statistical Learning and Modeling -- 36-600
subtitle: |
  Lab 1R: R Boot Camp\
  Due Thursday, August 28, 11:59pm\
author: Trishla Nair
output:
  pdf_document:
    toc: false
  html_document:
    toc: false
    code-overflow: wrap
    code-fold: false
    theme: spacelab
urlcolor: blue
---

If you're reading this: congratulations!  It means that you have successfully installed `R` and `RStudio` on your computer and that you are ready to dive into building the basic skills that will lead you to being able to manipulate data frames: tables of data where each row represents an observation (e.g., a person) and each column represents a measurement (e.g., the height or weight of that person).

Before we start, note that when you insert code into each "code chunk" below (the sections delimited with \`\`\`{r} and \`\`\`), you can test it in a number of ways:

- at the upper-right corner of a code chunk, there is a green triangle -- clicking on that runs all the code in the chunk
- also at the upper-right corner of a code chunk, there is a white triangle pointing downward toward a green line -- clicking that symbol runs all the code in previous chunks up to and **not** including the current one (attached to which you pressed the button; this means if you hit this button on the very first code chunk of a document, **nothing** will execute!)
- you can highlight any amount of code (and only code) and, on Windows, press Control+Enter, or on a Mac, press Command+Return, and the highlighted code will execute
- you can click on "Knit," which will run *all code in the file* and output an html file with the results


## Question 1

Each column of a data frame is itself a vector.  So let's start with some basic vector manipulation.  Use the `c()` function to define a vector `x` with four values: 1, 2, 1, and 8.  (*You should replace the "# FILL ME IN" statement below with your answer.*) Note that vectors are homogeneous (all of the same type), with the most common types being `double` (or `numeric`), `character`, and `logical`.  The vector you define has type `double`.  Check this by typing `typeof(x)` and noting the output.  Then type `x[1]`.  What do you see?  (Indicate your answer by, on a new line in the code chunk, typing one number symbol # [which denotes a comment] and then afterwards typing your answer in words [and/or numbers].)

```{r}
x <- c(1, 2, 1, 8)
x[1]
```

If the value(s) inside the square bracket is/are numeric, then that/those elements of the vector are displayed.  (*Note*: `R` counts from 1, not 0.)  If the value(s) are logical, then only those elements with value `TRUE` are displayed.  This will make more sense below.


## Question 2

Now define a vector `y` with four values: 2, 2, 5, 8.  Then add `x` and `y`, and multiply `x` and `y`.  Note that the following operators are using to carry out basic math in `R`:

| Operation | Description |
| --------- | ----------- |
| +         | addition |
| -         | subtraction |
| &ast;     | multiplication |
| /         | division |
| ^         | exponentiation |
| %%        | modulus (i.e., remainder) |
| %/%       | division with (floored) integer round-off |

```{r}
y <- c(2,2,5,8)
print(x+y)
print(x * y)
```

What you should observe are vectors with four numbers each.  Note that `R` did not require you to loop over the vector indices, i.e., `R` did not make you add `x[1]` and `y[1]` first, then `x[2]` and `y[2]`, etc.  `R` made things easy, by utilizing *vectorization*: it takes care of entire vectors at once, without explicit loops needing to be defined by you.


## Question 3

Now redefine the vector `x` to be of length 500, with all the elements being 0.  You don't want to do this using the `c()` function!  (Look back at section 2.1 of `cmu-intro-r.github.io` for alternative ways to define vectors.)

```{r}
x <- rep(c(0), each = 500)
length(x)
```


## Question 4

Let's define a random vector of integers with the `sample()` function.  The information that you pass into a function is called an argument, and `R` functions sometimes can have many arguments.  Let's look at the help page for `sample()`, which you can bring up by typing `?sample` in the console, or going to the `Help` pane and typing `sample` in the search bar.
```
Usage

sample(x, size, replace = FALSE, prob = NULL)
```

(What I just typed above is an example of a *verbatim* block: it **doesn't** execute as `R` code.)  What we see is that `sample()` has four arguments.  Two of them, `replace` and `prob`, have *default values*...so if you are happy with the defaults, you need not specify values for these arguments at all.  You **do** need to specify any arguments which have no default value.  Here, at a minimum, that would be two arguments: `x`, which is either a number or a vector from which to sample data, and `size`, which is the number of data to sample.  If you do this

```
x <- sample(10, 5)
```

you are telling `R` to sample five numbers between 1 and 10 (inclusive), with all the numbers being different (because `replace = FALSE`, by default), and to save the numbers as the vector `x`.  If you do this

```
x <- sample(40:50, 5)
```

you are telling `R` to sample five different numbers between 40 and 50 (inclusive, and without replacement).  And if you do this

```
x <- sample(3, 10, replace = TRUE)
```

you are telling `R` to sample ten numbers between 1 and 3 (inclusive), and now with repetition being allowed.  (We call this "sampling with replacement.")  Etc.  Now, sample 100 numbers between 1 and 100 (inclusive) with replacement, and save the output as the vector `x`.  How many unique integers are there in `x`?  Use handy vector functions (see section 2.3) to get a concise answer: do not print out `x` and count by eye!  (If you need help, call a TA or me over, or come to office hours.)

```{r}
x <- sample(1:100,size = 100,replace = TRUE)
unique(x)
```


## Question 5

Relational operators are binary operators of the form "<variable> <operator> <value>", e.g., `x < 0`.  The six basic relational operators are `==`, `!=`, `<`, `>`, `<=`, and `>=` (for "equals," "not equals," "less than," "greater than," "less than or equals," and "greater than or equals").  Relational operators return a vector of logicals, meaning a vector of `TRUE` and `FALSE` values.  Below, redefine `x` to be the vector with elements 1, 2, 1, and 8, and then display the output for `x == 1` and `x > 3`.

```{r}
x <- c(1,2,1,8)
print(x==1)
print(x > 3)
```


## Question 6

Apply the `sum()` function with input `x == 1`.  Does the output make sense to you?

```{r}
sum(x==1)
```


## Question 7

Relational operators may be combined with `&` (logical AND) or `|` (logical OR).  Below, display the output for `x < 2 | x > 5`.

```{r}
print(x<2 | x>5)
```


## Question 8

A reason to learn relational operators is that they underpin the manipulation of vectors (and thus underpin the manipulation of, e.g., rows or columns of data frames).  To display a subset of values of the vector `x`, you can for instance type `x[...]`, where you would replace `...` with a relational operator.  What happens when you type `x[x == 1]`?

```{r}
print(x[x==1])
```


## Question 9

Some last things to do for now: apply the `length()` function to `x`, apply the `sort()` function to `x`, apply the `sort()` function to `x` with the additional argument `decreasing = TRUE`, apply the `unique()` function to `x`, and apply the `table()` function to `x`.  (You may have done some similar things above when we told you to solve certain problems with handy vector functions.)  Try to form some intuition about what each does.  (*Note*: `table()` is a handy function for doing exploratory data analysis of categorical variables.)

```{r}
length(x)
sort(x, decreasing = TRUE)
unique(x)
```


## Question 10

(Looking ahead.)  A *list* in `R` is a collection of other objects, let's say of vectors, for now.  Define a list below using `list()`, with the first argument being a defined vector with name `x` and values 1 and 2, and the second argument being a defined vector with name `y` and values "a", "b", and "c". (*Note*: your arguments won't look like `z <- c(TRUE, FALSE)` but instead like `z = c(TRUE, FALSE)`).  Display the list.

```{r}
x <- c(1,2)
y <-c("a", "b", "c")
z <-list(x = x, y = y)
print(z)
```

The individual entries of a list are vectors, which are homogeneous, but the entries may each be of **different** type.  **A list whose entries are all of the same length is a data frame (i.e., a structured data table).**


## Question 11

Download `simple.txt` from the Week 1 module page.  Apply an appropriate function to read the file's contents into `R`.  (Are the data separated by commas or spaces?  Note that when you click on the file link, the file's contexts should appear in `Canvas`, along with a download link.)  Show the names of the columns.  Make sure the names are correct, and that there are eight columns.  (Use `ncol()` to determine the number of columns.)  *Note*: you should consider including the argument `stringsAsFactors = FALSE`.

```{r}
table <- read.table("C:/Users/Trish/Downloads/simple.txt", header = TRUE, stringsAsFactors = FALSE)
print(table)
```


## Question 12

Read in the data file from Question 11 but skip the header.  Display the names that `R` gives to the columns.  Note that the `skip` argument can be very useful for skipping over any metadata that may be present in your file.

```{r}
table <- read.table("C:/Users/Trish/Downloads/simple.txt", header = FALSE, stringsAsFactors = FALSE,skip = 1)
print(table)
```


## Question 13

Read in the data file from Question 11 but only read in the first four lines, while retaining the header.  Print the data frame.  (*Note*: in general, you can use `skip` and `nrows` to zero in on portions of a text file where the data actually reside.)

```{r}
table <- read.table("C:/Users/Trish/Downloads/simple.txt", header = TRUE, stringsAsFactors = FALSE, nrows = 4)
print(table)
```


## Question 14

Download `planets_small.csv` from the Week 1 module page.  Apply an appropriate function to read the file's contents into `R`.  Note that here, you have one column that should be rendered as character strings (`pl_hostname`, the first column), while the rest should be rendered as factor variables.  Thus you do not want to use the `stringsAsFactors` argument here, as it is too coarse.  You need to explicitly specify the types of each column, using the functional argument `colClasses`.  After reading the data in, pass your data frame into the function `summary()` and see what happens.

```{r}
csv <-read.csv("C:/Users/Trish/Downloads/planets_small.csv", header = TRUE, colClasses =c('character', 'factor', 'factor', 'factor'),nrows = 5)
print(csv)
```


## Question 15

Download `emline.csv` from the Week 1 module page.  Apply an appropriate function to read the file's contents into `R`.  When you are done, show the mean and median values of the `sfr` column.  *Hint*: if they are wildly different, you may need to adjust how you read in the data.  *Hint*: look for numbers that represent missing data, and use an appropriate argument to tell `R` that those numbers should be converted to `NA`.  *Note*: if any are converted to `NA`, you will have to pass an additional argument into the `mean()` and `median()` functions: `na.rm = TRUE`, which tells `R` to ignore the `NA`s in computations.

```{r}
df <-read.csv("C:/Users/Trish/Downloads/emline.csv", header = TRUE,na.strings = "NA")
print(df)
df$sfr[df$sfr == -9999] <-NA
print(mean(df$sfr, na.rm = TRUE))
print(median(df$sfr, na.rm = TRUE))
```


