---
title: Overview of Statistical Learning and Modeling -- 36-600
subtitle: |
  Lab 3T: Data Manipulation with dplyr\
  Due Wednesday, September 10, 11:59pm\
author: Trishla Nair
output:
  pdf_document:
    toc: false
  html_document:
    toc: false
    code-overflow: wrap
    code-fold: false
    theme: spacelab
urlcolor: blue
---


## Data

We'll begin by importing some astronomical data from a repository I've made to collect all the data we'll use throughout the course.  These data are stored in `.Rdata` format; such data are saved via `R`'s `save()` function and loaded via `R`'s `load()` function.  One wrinkle here: the data are stored on the web, so we also have to apply the `url()` function.

```{r}
load(url("https://github.com/FSKoerner/F25-36600-data/raw/refs/heads/main/Buzzard_DC1.Rdata"))
set.seed(101)
s  <- sample(nrow(df), 4000)
df <- df[s, -c(7:13)]
```

`df` is a data frame with 4000 rows and 7 columns; the first six are *predictor variables* and the seventh, `redshift`, is the *response variable*, in the context of supervised learning.  Redshift is a directly observable proxy for the distance of a galaxy from the Earth.  (After all, tape measures aren't going to help us here.)  In a statistical learning exercise, we might try to determine a statistical model that optimally associates the predictor variables with the `redshift`.


## Question 1

Apply the `dim()`, `nrow()`, `ncol()`, and `length()` functions to `df`, so as to build intuition about what these functions output.  (*Note*: we are not using `dplyr` yet, just base `R` functions here.)  Do you know why `length()` returns the value it does?  Ask me or the TA if you do not.  (But fill in an answer below regardless to reinforce the answer.)

```{r}
print(dim(df))
print(nrow(df))
print(ncol(df))
print(length(df))
```

```
a df is like a list of lists where there are 7 lists of 4000 elements each.
```


## Question 2

Display the names of each column of `df`.

```{r}
colnames(df)
```


---

Time for a digression.

A *magnitude* is a logarithmic measure of brightness that is calibrated such as to be approximately zero for the brightest stars in the night sky (Sirius, Vega, etc.).  Every five-unit *increase* in magnitude is a factor of 100 *decrease* in brightness.  So a magnitude of 20 means the object is 100 million times fainter than a star like Vega.

Magnitudes are generally measured in particular bands.  Imagine that you put a filter in front of a telescope that only lets photons of certain wavelengths pass through.  You can then assess the brightness of an object at just those wavelengths.  So `u` represents a magnitude determined at ultraviolet wavelengths, with `g`, `r`, and `i` representing green, red, and infrared.  (The `z` and `y` are a bit further into the infrared.  Those letters don't represent words.)

So the predictor data consists of six magnitudes spanning from the near-UV to the near-IR.

---


## Question 3

Use the base `R` `summary()` function to get a textual summary of the data frame.  Do you notice anything strange?  (Some values you might not expect?)

```{r}
summary(df)
```

```
The max for some columns are 99.
```


---

## dplyr

Below, we will practice using `dplyr` package functions to select rows and/or columns of a data frame.  `dplyr` is part of the `tidyverse`, and is rapidly becoming the most oft-used way of transforming data.  To learn more about "transformatory" functions, read through today's class notes, then through Chapter 5 of the online version of *R for Data Science* (see the syllabus for the link, or just Google for it).  In short, you can:

| action | function |
| ------ | -------- |
| pick features by name | `select()` |
| pick observations by value | `filter()` |
| pick observations by category | `group_by()` |
| create new features | `mutate()` |
| reorder rows | `arrange()` |
| collapse rows to summaries | `summarize()` |

A cool thing about `dplyr` is that you can use a piping operator (%&gt;%) to have the output of one function be the input to the next.  And you don't have to have only `dplyr` functions within the flow; for instance, you could pipe the first two rows of your data frame to the `head()` function (which just displays the first few rows of an object):

```{r}
suppressMessages(library(tidyverse))
head(df[,1:2])                           # base R
df %>% select(., u, g) %>% head(.)       # dplyr 
```

*Note*: if you've used `tidyverse` syntax before, you might notice that we don't always even need the `.`, which indicates where the previous object should be passed into the subsequent function.  In fact, we can exclude it entirely, but then `R` will assume the preceding output should go into the succeeding function's **first** argument.  If that's what you want, and you don't need to specify any other arguments, you can even ditch the parentheses entirely.  E.g., the following line does exactly the same thing as the last two: 

```{r}
df %>% select(u, g) %>% head()             # dplyr, but shorter
```

Now let's do a few exercises.  Ask us for extra background/detail!  (Or, computing problems tend to often be solved in the real world, tap into, e.g., StackOverflow and *R for Data Science*.)

---


## Question 4

Grab all data for which `i` is less than 25 and `g` is greater than 22, and output it in order of increasing `y`.  (Remember: you combine conditions with &amp; for "and" and | for "or".)  Show only the first six lines of output.  Note that `head()` by default shows the first six rows of the input data frame.

```{r}
df %>%arrange(., y)%>%filter(., (i<25 & g>22)) %>% head(., 6)
```

## Question 5

To get a quick, textual idea of how the data are distributed: select the `g` column, pipe it to the `round()` function, then pipe the output of `round()` to `table()`.  You should notice something strange about the output...the same thing you should have noticed when answering Question 3.

```{r}
df%>%select(.,g)%>%round()%>%table()
```


---

Time for another digression.  Domain scientists are not bound by the `R` convention of using `NA` when data are "not available," or missing.  Sometimes the domain scientists will tell you what they use in place of `NA`, sometimes not.  In those latter cases, one can usual infer the values.  Astronomers for some reason love using -99, -9, 99, etc., to represent missing values.  The values of 99 in the table you generated in Question 5 actually represent missing data.

We could change the values of 99 to `NA`, but here we will just filter those rows with values of 99 out of the data frame.

---


## Question 6

Use `filter()` to determine how many rows contain 99's.  Since 99's can appear in different columns, you will need to combine conditions together with logical "and"s or "or"s.  *Note*: only four of the six columns have 99's in them.

```{r}
df%>%filter(., (u ==99 | g == 99 | z == 99 | y == 99))%>%head()
```


## Question 7

Now repeat Question 6 (in a sense), and remove all the rows that contain 99's, saving the new data frame as `df.new`.  *Hint*: the opposite of, e.g., `u == 99 | g == 99` is `u != 99 & g != 99`, etc.  If you apply `dim()` to the new data frame, you should see that 3607 rows are retained.

```{r}
df.new <- df%>%filter(., (u !=99 & g != 99 & z != 99 & y != 99))
dim(df.new)
```


---

The data we are working with have no factor variables, so let's create one on the fly, using the `df.new` object you created in the last question (make sure to remove the chunk option `eval=FALSE`):

```{r}
type    <- rep("FAINT", nrow(df.new))
w       <- which(df.new$i < 25)
type[w] <- "BRIGHT"
type    <- factor(type)
unique(type)
df.new  <- cbind(type, df.new)
```

So, we've defined our factor variable using character strings, and then coerced the vector of strings into a factor variable with two levels.  Note that by default, the levels are ordered alphabetically.  For most factor variables, this wouldn't be, in any sense, a *natural* ordering.  You can override that default behavior if there actually is a natural ordering to any given factor variable.  See the documentation for `factor()` to see how to do that.

---


## Question 8

Use `group_by()` and `summarize()` to determine the numbers of `BRIGHT` and `FAINT` galaxies in the dataset.  (If you are adventuresome: try implementing the `tally()` function instead of `summarize()`.  For our purposes here, it's actually easier.)

```{r}
df.new%>%group_by(., type)%>%tally(.,)%>%as_tibble(.name_repair = "unique")
```


## Question 9

Repeat Question 8, but show the median value of the `u` magnitude instead of the counts in each factor group.  (You do need `summarize()` here; `tally()` won't help you.)

```{r}
df.new%>%group_by(., type)%>%summarize(Median = median(u))
```


---

Time for yet another digression.

Magnitudes of galaxies at particular wavelengths are heavily influenced by two factors: physics (what is going on with the gas, dust, and stars within the galaxy itself), and distance (the further away a galaxy is, the less bright it tends to be, so the magnitude generally goes up).  To attempt to mitigate (somewhat, but not necessarily completely) the effect of distance, astronomers often use *colors*, which are defined as differences in magnitude for two adjacent filters.

---


## Question 10

Use `mutate()` to define two new columns for `df.new`: `gr`, which would be the `g` magnitude minus the `r` magnitude, and `ri`, for `r` and `i`.  Save your result as `df.newer`.

```{r}
df.newer<- df.new%>%mutate(., gr = g-r)%>%mutate(., ri = r-i)
df.newer%>%head()
```


## Question 11

Are the mean values of `gr` and `ri` roughly the same for `BRIGHT` galaxies versus `FAINT` ones?  I haven't the faintest.  Use `dplyr` functions to attempt to answer this question.

```{r}
means_gr <-df.newer%>%group_by(., type)%>%summarize(Mean = mean(gr))
means_ri <-df.newer%>%group_by(., type)%>%summarize(Mean = mean(ri))
print(means_gr)
print(means_ri)
```


## Question 12

Let's go back to hypothesis testing for a moment.  Let's extract two vectors of data from `df.newer` (again, remove the chunk option `eval=FALSE`):

```{r}
gr.faint  <- df.newer$gr[df.newer$type == "FAINT"]
gr.bright <- df.newer$gr[df.newer$type == "BRIGHT"]
length(gr.faint)
length(gr.bright)
```

The first vector has all the `gr` colors for faint galaxies, and the second vector has all the ones for bright galaxies.  The code here *should* be familiar given what we covered in Week 1, but if you are not sure what this code is doing, call me or the TA over!

Here are two histograms, one for `gr.faint` and one for `gr.bright` (again, remove the chunk option `eval=FALSE`):

```{r eval=FALSE}
hist(gr.faint, prob = TRUE, main = NULL, col = alpha("firebrick", 0.4), 
     ylim = c(0, 1.2), xlab = "color", breaks = seq(-1.5, 5.5, by = 0.2))
hist(gr.bright, prob = TRUE, main = NULL, col = alpha("dodgerblue", 0.4), 
     breaks = seq(-1.5, 5.5, by = 0.2), add = TRUE)
```

Now: test the hypothesis that the means of the distributions that `gr.faint` and `gr.bright` are sampled from are the same.  Use a two-sample $t$-test.  What do you conclude?  Is a two-sample $t$-test actually strictly appropriate here?  How would you know?  (Even if the answer is no, do the $t$ test anyway.)

```{r}
print(t.test(gr.faint, gr.bright))
```

```
Because the calculated p-value is extremely small, there is a significant difference between the means of gr.faint and gr.bright. WHile a t-test can be used, a z test would work just fine.However, given the plot, the data appears to be skewed.
```

