---
title: "Project 1"
author: "Trishla Nair"
date: "2025-09-23"
output:
  pdf_document: default
  html_document: default
---
The data we are examining is from CMU's github. We are investigating if sleep patterns can affect GPA of freshman.
```{r}
library(ggplot2)
library(tidyverse)
library(corrplot)
library(RColorBrewer)

df <- read.csv("C:/Users/Trish/Documents/CMU/CMU-3660/Labs/cmu-sleep.csv")
df <- df%>% select(-subject_id)
print(head(df))
```

Firstly, I removed subject id as it had the least relelvance to what I wish to examine (that being the relationship between sleep and cummulative gpa). Then, using only quantitative data, I examined the data for outliers.
```{r}
numeric_cols <- sapply(df, is.numeric)
df_numeric <- df[, numeric_cols]

boxplot(df_numeric, main = "Box Plots",
        xlab = "Features", ylab = "Values", las = 2, cex.axis = 0.5, col = brewer.pal(n = 10, name = "Paired"))
```

 Based on the boxplots, it seems for the columns related to sleep, have the most outliers. So, next I factorized the demo_ columns and checked for rows with Nan values to remove them.
```{r}
df <- df %>%
  mutate(cohort = as.factor(cohort),
         demo_gender = as.factor(demo_gender), 
         demo_firstgen = as.factor(demo_firstgen), 
         demo_race = as.factor(demo_race), 
         cohort = as.factor(cohort))
```
Check for NaN's
```{r}
colSums(sapply(df, is.nan))
summary(df)
```
Remove Nan's
```{r}
df_cleaned <- df %>% drop_na()
summary(df_cleaned)
```
I then built a bar graph of mean sleep of each cohort per gender to see if there possibly wether the school or gender may yeild interesting insights on the average sleep a freshamn gets.
```{r}
#
pl<- df_cleaned %>%group_by(cohort, demo_gender) %>%
  summarize(mean_sleep = mean(TotalSleepTime, na.rm = TRUE), .groups = 'drop')

ggplot(pl, aes(x = cohort, y = mean_sleep, fill = demo_gender)) +
  geom_col(position = position_dodge(width = 1)) +
  geom_text(aes(label = round(mean_sleep, 3)),
            position = position_dodge(width = 1),
            vjust = -0.5) +
  scale_fill_brewer(palette = "Pastel2") +
  labs(title = "Mean sleep time per cohort based on Gender",x = "Cohort", y = "Mean Sleep Time", fill = "Gender")
```
I then constructed a paired plot examining total sleep and midpoint sleep per cohort to see if the school plays a factor into wether sleep affects gpa.
```{r}
sleep <- df_cleaned %>%
  pivot_longer(cols = c(TotalSleepTime, midpoint_sleep),
               names_to = "Measure",
               values_to = "Value")
ggplot(sleep, aes(x = Measure, y = Value, group = cohort, color = cohort)) +
  geom_line(alpha = 0.5) +
  geom_point(size = 2) +
  facet_wrap(~ cohort) +   # <-- split into panels
  labs(title = "Paired Plot of Total Sleep vs Midpoint Sleep per Cohort",
       x = "Sleep features",
       y = "Length in minutes")
```
Based on the pair plots, the UW cohorts seem to sleep longer compared to the LAC cohorts, as indicated by the fact the LAC cohorts seems to have midpoint_sleep time much later than the UW students. I then did a chi-square test to examine the possible correlation between mean sleep time and mean gpa.
```{r}
cor.test(df_cleaned$TotalSleepTime, df_cleaned$term_gpa, use = "complete.obs")
lm_gpa <- lm(term_gpa ~ TotalSleepTime, data = df_cleaned)
summary(lm_gpa)
```
Based on my findings, since the p value is so small, we reject the null hypothesis and see that there is a significant correlation between the 2. Thus, I created a correlation matrix to see what other variables might affect term_gpa.
```{r}
num_cols <- sapply(df_cleaned, is.numeric)
df_num <- df_cleaned[, num_cols]
matrix<-cor(df_num)
corrplot(matrix, method="number", number.cex = 0.5, col = brewer.pal(n = 8, name = "Spectral"))
```
After running a correlation matrix, the only feature that may possess a strong correlation with term GPA is the cumulative GPA. I then ran a hypothesis test to examine whether the time may effect the term gpa by comparing the first and fifth studies at a 95% confidence level.
```{r}
s_1 <- df_cleaned%>%select(term_gpa, study)%>%filter(study == 1)
s_5 <- df_cleaned%>%select(term_gpa, study)%>%filter(study == 5)
t_test_result <- t.test(s_1$term_gpa, s_5$term_gpa,alternative = "less", var.equal = FALSE, conf.level = 0.95)

print(t_test_result)

```
Given how large the pvalue is, we fail to reject the null hypothesis and thus there is no difference in term GPA's between studies 1 and 5.
